Baseline Drift Estimation for Air Quality Data Using Quantile Trend Filtering

### 2019-04-11 Manuscript/detrendify_aoas.tex
removed \mathcal{I} notation, added note on differences between objectives, and corrected the theta update statement

We combine the ideas of quantile regression and trend filtering. For a single quantile level $\tau$, the quantile trend filtering problem is given in \Eqn{quantile_trend}. As with classic quantile regression, the quantile trend filtering problem is a linear program which can be solved by a number of methods. We want to estimate multiple quantiles simultaneously and to ensure that our quantile estimates are valid by enforcing the constraint that if $\tau_2 > \tau_1$ then $Q(\tau_2) \ge Q(\tau_1)$ where $Q$ is the quantile function of $\V{y}$. Even if a single quantile is ultimately desired, ensuring non-crossing allows information from nearby quantiles to be used to improve the estimates. Given quantiles $\tau_1 < \tau_2 < \ldots < \tau_J$, the optimization problem becomes
\begin{eqnarray*}
	\label{eq:noncross_trend}
	&&\underset{\M{\Theta} \in \mathcal{C}}{\text{minimize}}\; \sum_{j=1}^J \left [\rho_{\tau_j}(\V{y} - \V{\theta}_{j}) +
	\lambda_j \lVert \Mn{D}{k+1} \V{\theta}_j \rVert_1 \right ] \\
\end{eqnarray*}
where $\M{\Theta} = \begin{pmatrix} \V{\theta}_1 & \V{\theta}_2 & \cdots & \V{\theta}_J \end{pmatrix} \in \Real^{n \times J}$ is a matrix whose $j$th column corresponds to the $j$th quantile signal $\V{\theta}_j$ and the set $\mathcal{C} = \{\M{\Theta} \in \Real^{n \times J}: \ME{\theta}{ij} \leq \ME{\theta}{ij'} \;\text{for $j \leq j'$}\}$ encodes the non-crossing quantile constraints. The additional constraints are linear in the parameters, so the non-crossing quantile trends can still be estimated by a number of available solvers. We allow for the possibility that the degree of smoothness in the trends varies by quantile by allowing the smoothing parameter to vary with quantile as well. In the rest of this paper, we report numerical results using the commercial solver Gurobi \citep{gurobi} and its R package implementation. However, we could easily substitute a free solver such as the Rglpk package by \cite{rglpk}.

@@

 Let $\mathcal{I}$ denote a subset of the indices $\{1, \ldots, n\}$ and $\V{y}_\mathcal{I} \in \Real^{\lvert I \rvert}$ denote the subvector obtained by stacking on top of each other the elements $\VE{y}{i}$ for $i \in \mathcal{I}$ in their canonical ordering.
 To decrease computation time and extend our method to larger problems, we divide our observed series $\VE{y}{i}$ with $i = 1, \ldots, n$ into $W$ overlapping windows of observations $\Vn{y}{w} = \V{y}_{\mathcal{I}_w}$ where $\mathcal{I}_w$ is the set of integers from $l_w$ to $u_w$ inclusive, namely $\mathcal{I}_w = \{l_w, l_w + 1, \ldots, u_w -1, u_w\}$ and
 \begin{eqnarray*}
 	1 = l_{1} < l_{2} < u_{1} < l_{3} < u_{2} < l_{4} < u_{3} < \cdots < u_{W} = n.
 \end{eqnarray*}
 \Fig{windows} shows an example of 1200 observations being mapped into three equally sized overlapping windows of observations. While the overlapping trend estimates between $l_2$ and $u_2$ don't vary dramatically, the difference is more pronounced in the trend in the 5\textsuperscript{th} quantile between $l_3$ and $u_2$.

 Given quantiles $\tau_1 < \cdots < \tau_J$, we introduce dummy variables $\Vn{\theta}{w}_{j} \in \Real^{\lvert \mathcal{I}_w \rvert}$ as the value of the $\tau_j$\textsuperscript{th} quantile trend in window $w$. We then ``stitch" together the $W$ quantile trend estimates into consensus over the overlapping regions by introducing the constraint $\VnE{\theta}{w}{ij} = \VE{\theta}{i + l_w - 1,j}$ for $i = 1, \ldots, u_w - l_w + 1$ and for all $j$. Let $\Mn{\Theta}{w}$ be the matrix whose $j$th column is $\Vn{\theta}{w}_{j}$. Then we can write these constraints more compactly as $\Mn{\Theta}{w} = \Mn{U}{w}\M{\Theta}$, where $\Mn{U}{w} \in \{0, 1\}^{\lvert \mathcal{I}_w \rvert \times n}$ is a matrix that selects rows of $\M{\Theta}$ corresponding to the $w$th window, namely
 \begin{eqnarray*}
 	\Mn{U}{w} & = & \begin{pmatrix}
 		\V{e}_{l_w}\Tra \\
 		\vdots \\
 		\V{e}_{u_w}\Tra
 	\end{pmatrix},
 \end{eqnarray*}
 where $\V{e}_i \in \Real^{n}$ denotes the $i$th standard basis vector. Furthermore, let $\iota_{\mathcal{C}}$ denote the indicator function of the non-crossing quantile constraint, namely $\iota_{\mathcal{C}}(\M{\Theta})$ is zero if $\M{\Theta} \in \mathcal{C}$ and infinity otherwise.

@@

 The ADMM algorithm \citep{gabay1975dual, glowinski1975approximation} is  described in greater detail by \cite{boyd2011distributed}, but we briefly review how it can be used to iteratively solve the following equality constrained optimization problem which is a more general form of \Eqn{quantile_windows}.

@@

{\bf Updating $\{\Mn{\Theta}{w}\}$: } We then estimate the trend separately in each window, which can be done in parallel, while constraining the overlapping pieces of the trends to be equal as outlined in \Alg{admm}. The use of the Augmented Lagrangian converts the problem of solving a potentially large linear program into a solving a collection of smaller quadratic programs. The \texttt{gurobi} R package \citep{gurobi} can solve quadratic programs in addition to linear programs, but we can also use the free R package \texttt{quadprog} \citep{quadprog}.

>>>>

We combine the ideas of quantile regression and trend filtering. For a single quantile level $\tau$, the quantile trend filtering problem is given in \Eqn{quantile_trend}. As with classic quantile regression, the quantile trend filtering problem is a linear program which can be solved by a number of methods. We want to estimate multiple quantiles simultaneously and to ensure that our quantile estimates are valid by enforcing the constraint that if $\tau_2 > \tau_1$ then $Q(\tau_2) \ge Q(\tau_1)$ where $Q$ is the quantile function of $\V{y}$. Even if a single quantile is ultimately desired, ensuring non-crossing allows information from nearby quantiles to be used to improve the estimates. Given quantiles $\tau_1 < \tau_2 < \ldots < \tau_J$, the optimization problem becomes
\begin{eqnarray}
	\label{eq:noncross_trend}
	&&\underset{\M{\Theta} \in \mathcal{C}}{\text{minimize}}\; \sum_{j=1}^J \left [\rho_{\tau_j}(\V{y} - \V{\theta}_{j}) +
	\lambda_j \lVert \Mn{D}{k+1} \V{\theta}_j \rVert_1 \right ] 
\end{eqnarray}
where $\M{\Theta} = \begin{pmatrix} \V{\theta}_1 & \V{\theta}_2 & \cdots & \V{\theta}_J \end{pmatrix} \in \Real^{n \times J}$ is a matrix whose $j$th column corresponds to the $j$th quantile signal $\V{\theta}_j$ and the set $\mathcal{C} = \{\M{\Theta} \in \Real^{n \times J}: \ME{\theta}{ij} \leq \ME{\theta}{ij'} \;\text{for $j \leq j'$}\}$ encodes the non-crossing quantile constraints. The additional constraints are linear in the parameters, so the non-crossing quantile trends can still be estimated by a number of available solvers. We allow for the possibility that the degree of smoothness in the trends varies by quantile by allowing the smoothing parameter to vary with quantile as well. In the rest of this paper, we report numerical results using the commercial solver Gurobi \citep{gurobi} and its R package implementation. However, we could easily substitute a free solver such as the Rglpk package by \cite{rglpk}.

@@

 To decrease computation time and extend our method to larger problems, we divide our observed series $\VE{y}{i}$ with $i = 1, \ldots, n$ into $W$ overlapping windows of observations, defining the vector of sequential elements indexed from $l_w$ to $u_w$ as $\Vn{y}{w} = \{y_{l_w}, y_{l_w + 1}, \ldots, y_{u_w -1}, y_{u_w}\}$, with 
 \begin{eqnarray*}
 	1 = l_{1} < l_{2} < u_{1} < l_{3} < u_{2} < l_{4} < u_{3} < \cdots < u_{W} = n.
 \end{eqnarray*}
 We define $n_w = u_w-l_w+1$ so that $\Vn{y}{w} \in \Real^{n_w}$. \Fig{windows} shows an example of 1200 observations being mapped into three equally sized overlapping windows of observations. While the overlapping trend estimates between $l_2$ and $u_2$ don't vary dramatically, the difference is more pronounced in the trend in the 5\textsuperscript{th} quantile between $l_3$ and $u_2$.

 Given quantiles $\tau_1 < \cdots < \tau_J$, we introduce dummy variables $\Vn{\theta}{w}_{j} \in \Real^{n_w}$ as the value of the $\tau_j$\textsuperscript{th} quantile trend in window $w$. We then ``stitch" together the $W$ quantile trend estimates into consensus over the overlapping regions by introducing the constraint $\VnE{\theta}{w}{ij} = \VE{\theta}{i + l_w - 1,j}$ for $i = 1, \ldots, u_w - l_w + 1$ and for all $j$. Let $\Mn{\Theta}{w}$ be the matrix whose $j$th column is $\Vn{\theta}{w}_{j}$. Then we can write these constraints more compactly as $\Mn{\Theta}{w} = \Mn{U}{w}\M{\Theta}$, where $\Mn{U}{w} \in \{0, 1\}^{n_w \times n}$ is a matrix that selects rows of $\M{\Theta}$ corresponding to the $w$th window, namely
 \begin{eqnarray*}
 	\Mn{U}{w} & = & \begin{pmatrix}
 		\V{e}_{l_w}\Tra \\
 		\vdots \\
 		\V{e}_{u_w}\Tra
 	\end{pmatrix},
 \end{eqnarray*}
 where $\V{e}_i \in \Real^{n}$ denotes the $i$th standard basis vector. Furthermore, let $\iota_{\mathcal{C}}$ denote the indicator function of the non-crossing quantile constraint, namely $\iota_{\mathcal{C}}(\M{\Theta})$ is zero if $\M{\Theta} \in \mathcal{C}$ and infinity otherwise.

@@

 The solution to \Eqn{quantile_windows} is {\em not} identical to the solution to \Eqn{noncross_trend} because of double counting of the overlapping sections. The solutions are very close, however, and the differences are essentially immaterial concerning downstream analysis. The ADMM algorithm \citep{gabay1975dual, glowinski1975approximation} is  described in greater detail by \cite{boyd2011distributed}, but we briefly review how it can be used to iteratively solve the following equality constrained optimization problem which is a more general form of \Eqn{quantile_windows}.

@@

{\bf Updating $\{\Mn{\Theta}{w}\}$: } We then estimate the trend separately in each window, which can be done in parallel, while penalizing the differences in the overlapping pieces of the trends  as outlined in \Alg{admm}. The use of the Augmented Lagrangian converts the problem of solving a potentially large linear program into a solving a collection of smaller quadratic programs. The \texttt{gurobi} R package \citep{gurobi} can solve quadratic programs in addition to linear programs, but we can also use the free R package \texttt{quadprog} \citep{quadprog}.
